---
title: "5. Linear Mixed-Effects Model"
subtitle: "Accounting for block variation in an agricultural experiment"
author: "Gustavo Roa"
format:
  html:
    self-contained: true
editor: source
---

# Linear Mixed-Effects Models

Agricultural experiments often use blocks to account for variation across the field. A **linear mixed-effects model (LMM)** allows us to include treatment effects as fixed effects and block effects as random effects.

In this example, we will ask:

> **Does grain yield differ among sulfur rates after accounting for variation among blocks?**

The model is:

$$
Y_{ij} = \mu + \tau_i + b_j + \varepsilon_{ij}
$$

where:

- $Y_{ij}$ is the observed grain yield.
- $\mu$ is the overall mean.
- $\tau_i$ is the fixed effect of sulfur rate.
- $b_j$ is the random effect of block.
- $\varepsilon_{ij}$ is the residual error.

# Fixed and random effects

A **fixed effect** is a factor whose specific levels are of scientific interest.

In this example:

- **Sulfur rate** is a fixed effect because we want to compare the tested rates.
- **Block** is a random effect because we want to account for variation among blocks.

The choice of fixed or random effects depends on the research question and experimental design. For example, location or year could be treated differently in another study.

# Load the required packages

```{r, warning=FALSE, message=FALSE}
# Import Excel files
library(readxl)

# Select and manipulate variables
library(dplyr)

# Create graphs
library(ggplot2)

# Fit linear mixed-effects models
library(lme4)

# Obtain Kenward-Roger tests
library(lmerTest)

# Estimate marginal means
library(emmeans)

# Create compact letter displays
library(multcomp)
library(multcompView)
```

# Load the data

We will use the same dataset used in the previous sections.

```{r}
# Import the workshop dataset
AllData <- read_excel(
  "Data_example.xlsx",
  sheet = "data"
)

# Check the structure of the dataset
str(AllData)
```

# Prepare the variables

Sulfur rate is converted to a factor because we want to compare the specific rates used in the experiment.

Block is converted to a factor because it identifies the experimental blocks.

```{r}
# Convert sulfur rate to a treatment factor
AllData$S_rate_kg_ha <- factor(AllData$S_rate_kg_ha)

# Convert block to a grouping factor
AllData$block <- factor(AllData$block)
```

# Fit the mixed model

We can fit the model using `lmer()`.

The model is:

```r
grain_yield_Mg_ha ~ S_rate_kg_ha + (1 | block)
```

Here:

- `grain_yield_Mg_ha` is the response variable.
- `S_rate_kg_ha` is the fixed effect.
- `(1 | block)` specifies block as a random effect.

```{r}
# Fit the linear mixed-effects model
model <- lmer(
  grain_yield_Mg_ha ~ S_rate_kg_ha + (1 | block),
  data = AllData,
  REML = TRUE
)

# Display the model summary
summary(model)
```

# Test the sulfur effect

For mixed models, the denominator degrees of freedom need to be estimated.

The **Kenward-Roger (KR) method** provides an adjustment for these degrees of freedom when testing fixed effects.

```{r}
# Test the fixed effect using Kenward-Roger degrees of freedom
anova(
  model,
  ddf = "Kenward-Roger"
)
```

The ANOVA table includes:

- **NumDF:** numerator degrees of freedom.
- **DenDF:** denominator degrees of freedom.
- **F value:** test statistic.
- **Pr(>F):** P-value for the sulfur effect.

The main question is:

> **Is there evidence that grain yield differs among sulfur rates after accounting for block variation?**

# Estimated marginal means

After testing the treatment effect, we can estimate the mean grain yield for each sulfur rate.

The `emmeans()` function calculates **estimated marginal means** from the fitted model.

```{r}
# Estimate marginal means using Kenward-Roger degrees of freedom
means <- emmeans(
  model,
  ~ S_rate_kg_ha,
  lmer.df = "kenward-roger"
)

# Display the estimated means
means
```

# Mean separation

When a treatment effect is of interest, we may want to compare individual treatment means.

There are several methods for multiple comparisons, including:

- Tukey
- Sidak
- Bonferroni
- Dunnett
- Fisher's LSD

The appropriate method depends on the comparisons planned and the research question.

Here, we will use **Sidak-adjusted comparisons** and summarize them with a compact letter display.

```{r}
# Create compact letter display
results <- cld(
  means,
  Letters = letters,
  lmer.df = "kenward-roger",
  sort = FALSE
) %>%
  as.data.frame() %>%
  mutate(
    letter = trimws(.group)
  )

# Display the treatment means and letters
results
```

The `letter` column summarizes the pairwise comparisons.

Treatments that share a letter are not significantly different according to the selected comparison procedure.

# Plot the estimated means

We can visualize the estimated means, confidence intervals, and letter groups.

```{r, echo=FALSE}
# Plot estimated means with 95% confidence intervals
ggplot(
  results,
  aes(
    x = S_rate_kg_ha,
    y = emmean
  )
) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(
      ymin = lower.CL,
      ymax = upper.CL
    ),
    width = 0.15
  ) +
  geom_text(
    aes(
      y = upper.CL,
      label = letter
    ),
    vjust = -0.5,
    size = 5
  ) +
  labs(
    x = "Sulfur rate (kg S/ha)",
    y = "Estimated grain yield (Mg/ha)"
  ) +
  theme_bw()
```

The figure shows:

- Estimated grain yield for each sulfur rate.
- 95% confidence intervals.
- Statistical groupings from the pairwise comparisons.

# Model assumptions

The main assumptions are:

- Observations are independent according to the experimental design.
- Residuals are approximately normally distributed.
- Residual variance is reasonably constant.

A simple residual plot can help identify obvious problems.

```{r, echo=FALSE}
# Create fitted values and residuals
diagnostic_data <- data.frame(
  fitted = fitted(model),
  residual = resid(model)
)

# Plot residuals against fitted values
ggplot(
  diagnostic_data,
  aes(
    x = fitted,
    y = residual
  )
) +
  geom_point() +
  geom_hline(
    yintercept = 0,
    linetype = "dashed"
  ) +
  labs(
    x = "Fitted values",
    y = "Residuals"
  ) +
  theme_bw()
```

We are looking for residuals that are reasonably scattered around zero without strong patterns.

# Why use the mixed model?

The model:

```r
grain_yield_Mg_ha ~ S_rate_kg_ha + (1 | block)
```

accounts for two main sources of variation:

- **Sulfur rate:** the fixed effect we want to study.
- **Block:** variation associated with the experimental layout.

This allows the treatment comparison to account for the blocked design.

# Key takeaways

- A mixed model includes fixed and random effects.
- Sulfur rate is a fixed effect because we want to compare the tested rates.
- Block is a random effect because it accounts for variation among blocks.
- `(1 | block)` specifies a random intercept for block.
- `lmer()` fits the mixed model.
- The Kenward-Roger method provides adjusted degrees of freedom for testing fixed effects.
- `emmeans()` estimates treatment means from the fitted model.
- Multiple-comparison methods can be used to compare treatment means.
- `cld()` summarizes pairwise comparisons using letters.
- Residual diagnostics help evaluate model assumptions.
- The experimental design should guide the model structure.

> **The statistical model should reflect the experimental design. In a blocked agricultural experiment, block variation can be incorporated as a random effect.**

