---
title: "6. Nonlinear Regression"
subtitle: "Modeling crop response to fertilizer rates"
author: "Gustavo Roa"
format:
  html:
    self-contained: true
editor: source
---

# Nonlinear Regression

Regression models describe the relationship between a response variable and one or more explanatory variables.

A **linear regression** describes a relationship using a straight line:

$$
Y_i = \beta_0 + \beta_1X_i + \varepsilon_i
$$

where:

- $Y_i$ is the response variable.
- $X_i$ is the explanatory variable.
- $\beta_0$ is the intercept.
- $\beta_1$ is the slope.
- $\varepsilon_i$ is the random error.

A **nonlinear regression** is useful when a straight line does not adequately describe the relationship between the response and explanatory variable.

Examples of nonlinear response models include:

- **Quadratic:** describes a curved relationship.
- **Exponential:** describes a response that changes at a changing rate.
- **Linear-plateau:** describes an increase followed by a plateau.
- **Quadratic-plateau:** describes a curved response followed by a plateau.

In agronomy, nonlinear models are commonly useful for describing crop responses to fertilizer rates.

In this section, we will focus on two models:

- Linear-plateau (LP)
- Quadratic-plateau (QP)

# Linear-plateau model

The **linear-plateau (LP)** model assumes that yield increases linearly until a breakpoint and then remains constant.

$$
Y =
\begin{cases}
\beta_0 + \beta_1X, & X \leq X_0 \\
Y_0, & X > X_0
\end{cases}
$$

where:

- $Y$ is grain yield.
- $X$ is sulfur rate.
- $\beta_0$ is the intercept.
- $\beta_1$ is the slope before the plateau.
- $X_0$ is the breakpoint.
- $Y_0$ is the plateau yield.

The **breakpoint** represents the estimated sulfur rate where the response reaches the plateau.

# Quadratic-plateau model

The **quadratic-plateau (QP)** model allows the response to curve before reaching the plateau.

$$
Y =
\begin{cases}
\beta_0 + \beta_1X + \beta_2X^2, & X \leq X_0 \\
Y_0, & X > X_0
\end{cases}
$$

where $\beta_2$ controls the curvature of the response.

The QP model can be useful when yield increases rapidly at lower fertilizer rates and the increase becomes smaller as fertilizer rate increases.

# Load the required packages

```{r, warning=FALSE, message=FALSE}
# Import Excel data
library(readxl)

# Create graphs
library(ggplot2)

# Fit nonlinear models
library(nlraa)

```

# Load the data

We will use the same dataset used in the previous workshop sections.

```{r}
# Import the workshop dataset
AllData <- read_excel(
  "Data_example.xlsx",
  sheet = "data"
)

# Check the data structure
str(AllData)
```

# Explore the response

Before fitting a nonlinear model, it is useful to look at the relationship between sulfur rate and grain yield.

```{r, echo=FALSE}
# Plot grain yield against sulfur rate
ggplot(
  AllData,
  aes(
    x = S_rate_kg_ha,
    y = grain_yield_Mg_ha
  )
) +
  geom_point(size = 2.5) +
  labs(
    x = "Sulfur rate (kg S/ha)",
    y = "Grain yield (Mg/ha)"
  ) +
  theme_bw()
```

The shape of the relationship provides an initial indication of whether a nonlinear response model may be appropriate.

# Fit the nonlinear models

## Linear-plateau

```{r}
# Fit the linear-plateau model
model_LP <- nls(
  grain_yield_Mg_ha ~ SSlinp(
    S_rate_kg_ha,
    a,
    b,
    xs
  ),
  data = AllData
)

# Display model results
summary(model_LP)
```

The parameter `xs` is the estimated breakpoint.

## Quadratic-plateau

```{r}
# Fit the quadratic-plateau model
model_QP <- nls(
  grain_yield_Mg_ha ~ SSquadp3xs(
    S_rate_kg_ha,
    a,
    b,
    xs
  ),
  data = AllData
)

# Display model results
summary(model_QP)
```

The QP model allows the response to curve before reaching the plateau.

# Compare the models

AIC can be used to compare models fitted to the same data.

$$
AIC = -2\log(L) + 2K
$$

where:

- $L$ is the likelihood of the model.
- $K$ is the number of estimated parameters.

Lower AIC indicates greater support among the models being compared.

```{r}
# Compare the LP and QP models using AIC
AIC(
  model_LP,
  model_QP
)
```

AIC is one tool for model comparison. The shape of the response and biological interpretation should also be considered.

# Plot the fitted models

We can plot both fitted models together with the observed data.

```{r, echo=FALSE}
# Create sulfur rates for prediction
new_data <- data.frame(
  S_rate_kg_ha = seq(
    min(AllData$S_rate_kg_ha),
    max(AllData$S_rate_kg_ha),
    length.out = 50
  )
)

# Predict grain yield from each model
new_data$LP <- predict(
  model_LP,
  newdata = new_data
)

new_data$QP <- predict(
  model_QP,
  newdata = new_data
)

# Plot observations and fitted curves
ggplot() +
  geom_point(
    data = AllData,
    aes(
      x = S_rate_kg_ha,
      y = grain_yield_Mg_ha
    ),
    size = 2.5
  ) +
  geom_line(
    data = new_data,
    aes(
      x = S_rate_kg_ha,
      y = LP,
      linetype = "Linear-plateau"
    ),
    linewidth = 1
  ) +
  geom_line(
    data = new_data,
    aes(
      x = S_rate_kg_ha,
      y = QP,
      linetype = "Quadratic-plateau"
    ),
    linewidth = 1
  ) +
  labs(
    x = "Sulfur rate (kg S/ha)",
    y = "Grain yield (Mg/ha)",
    linetype = "Model"
  ) +
  theme_bw()
```

The fitted curves allow us to visually compare how the LP and QP models describe the observed response.

# Model averaging

Sometimes more than one model provides a reasonable description of the data. **Model averaging** is an approach that combines information from multiple candidate models rather than selecting only one.

We will not perform model averaging in this workshop. 

[Model averaging for estimation of optimum rate and critical values under model uncertainty](https://hdl.handle.net/2097/47202)

# Exercise

## Exercise 6.1: Fit the linear-plateau model

Fit a linear-plateau model using:

- `grain_yield_Mg_ha` as the response.
- `S_rate_kg_ha` as the explanatory variable.

```r
# Fit the linear-plateau model
model_LP_ex <- nls(
  ____________ ~ SSlinp(
    ____________,
    a,
    b,
    xs
  ),
  data = AllData
)

summary(model_LP_ex)
```

::: {.callout-tip collapse="true"}
## Show solution

```{r}
# Fit the linear-plateau model
model_LP_ex <- nls(
  grain_yield_Mg_ha ~ SSlinp(
    S_rate_kg_ha,
    a,
    b,
    xs
  ),
  data = AllData
)

summary(model_LP_ex)
```

:::

## Exercise 6.2: Fit the quadratic-plateau model

Fit a quadratic-plateau model using the same variables.

```r
# Fit the quadratic-plateau model
model_QP_ex <- nls(
  ____________ ~ SSquadp3xs(
    ____________,
    a,
    b,
    xs
  ),
  data = AllData
)

summary(model_QP_ex)
```

::: {.callout-tip collapse="true"}
## Show solution

```{r}
# Fit the quadratic-plateau model
model_QP_ex <- nls(
  grain_yield_Mg_ha ~ SSquadp3xs(
    S_rate_kg_ha,
    a,
    b,
    xs
  ),
  data = AllData
)

summary(model_QP_ex)
```

:::

## Exercise 6.3: Compare the models

Use AIC to compare the LP and QP models.

```r
# Compare the models using AIC
AIC(
  ____________,
  ____________
)
```

::: {.callout-tip collapse="true"}
## Show solution

```{r}
# Compare the models using AIC
AIC(
  model_LP_ex,
  model_QP_ex
)
```

:::

## Exercise 6.4: Interpret the results

Use the model summaries, AIC results, and fitted curves to answer:

1. Which model has the lower AIC?
2. What is the estimated breakpoint for each model?
3. How do the fitted curves differ?
4. Which model would you use to describe the response, and why?


> **Important:** AIC helps compare competing models, but model selection should also consider the shape of the response and its biological meaning.

# Key takeaways

- Nonlinear models describe relationships that are not adequately represented by a straight line.
- LP describes a linear increase followed by a plateau.
- QP allows a curved response before reaching a plateau.
- The breakpoint is an important parameter in fertilizer response models.
- AIC can be used to compare competing models.
- Model selection should consider both statistical evidence and biological meaning.

