---
title: "Workshop Exercises"
subtitle: "Applied Statistics in R: From Data to Results"
format:
  html:
    self-contained: true
---

# Load packages

```{r}
# Load packages
library(tidyverse)
library(readxl)
library(lme4)
library(lmerTest)
library(emmeans)
library(multcomp)
library(nlraa)
```

# Import the data

```{r}
# Import the workshop dataset
AllData <- read_excel(
  "Data_example.xlsx",
  sheet = "data"
)
```

# Section 1: Data Exploration

## Exercise 1.1: Numerical summary

Use `summary()` to describe `grain_yield_Mg_ha`.

```{r}
# Summarize grain yield
summary(______________)
```

## Exercise 1.2: Histogram

Create a histogram of grain yield.

```{r}
# Create a histogram
ggplot(
  AllData,
  aes(x = ____________)
) +
  geom_histogram(bins = ___) +
  labs(
    x = "____________",
    y = "____________"
  ) +
  theme_bw()
```

## Exercise 1.3: Box plot

Create a box plot comparing grain yield among sulfur rates.

```{r}
# Create a box plot
ggplot(
  AllData,
  aes(
    x = ____________,
    y = ____________
  )
) +
  geom_boxplot() +
  labs(
    x = "____________",
    y = "____________"
  ) +
  theme_bw()
```

## Exercise 1.4: Interpret the results

Answer:

1. What is the typical grain yield?
2. How much variation is present?
3. Which sulfur rate has the highest median yield?
4. Are there any potentially unusual observations?

# Section 2: Correlation

## Exercise 2.1: Calculate a correlation

Calculate the Pearson correlation between `N_conc_R1_pct` and `S_conc_R1_pct`.

```{r}
# Calculate the Pearson correlation
cor(
  AllData$____________,
  AllData$____________,
  method = "pearson"
)
```

## Exercise 2.2: Create a scatterplot

```{r}
# Create a scatterplot
ggplot(
  AllData,
  aes(
    x = ____________,
    y = ____________
  )
) +
  geom_point() +
  labs(
    x = "____________",
    y = "____________"
  ) +
  theme_bw()
```

## Exercise 2.3: Interpret the relationship

Answer:

1. Is the relationship positive or negative?
2. How strong is the association?
3. Does the relationship appear linear?

# Section 3: Linear Regression

## Exercise 3.1: Fit a linear regression

Use `N_conc_R1_pct` to predict `S_conc_R1_pct`.

```{r}
# Fit a linear regression
model_ex <- lm(
  ____________ ~ ____________,
  data = AllData
)

summary(model_ex)
```

## Exercise 3.2: Add the regression line

```{r}
# Plot the data and fitted regression line
ggplot(
  AllData,
  aes(
    x = ____________,
    y = ____________
  )
) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = TRUE
  ) +
  labs(
    x = "____________",
    y = "____________"
  ) +
  theme_bw()
```

## Exercise 3.3: Interpret the model

Answer:

1. What does the slope tell us?
2. What is the P-value for the slope?
3. What does $R^2$ tell us?
4. Does the model provide evidence of a linear relationship?

# Section 4: ANOVA

## Exercise 4.1: Fit the ANOVA

Use grain yield as the response and sulfur rate as the treatment.

```{r}
# Convert sulfur rate to a treatment factor
AllData$S_rate_kg_ha <- factor(____________)

# Fit the one-way ANOVA
model_ex <- aov(
  ____________ ~ ____________,
  data = AllData
)

summary(model_ex)
```

## Exercise 4.2: Compare treatment means

Use `emmeans()` and Tukey's method.

```{r}
# Estimate treatment means
emm_ex <- emmeans(
  model_ex,
  ~ ____________
)

# Compare treatment pairs
pairs(
  emm_ex,
  adjust = "tukey"
)
```

## Exercise 4.3: Interpret the results

Answer:

1. Is there evidence that grain yield differs among sulfur rates?
2. What is the ANOVA P-value?
3. Which treatment pairs differ?
4. Which sulfur rate has the highest estimated mean?

# Section 5: Linear Mixed Model

## Exercise 5.1: Fit a linear mixed model

Treat sulfur rate as a fixed effect and block as a random effect.

```{r}
# Fit the linear mixed model
model_ex <- lmer(
  ____________ ~ ____________ + (1 | ____________),
  data = AllData,
  REML = TRUE
)

# ANOVA using Kenward-Roger degrees of freedom
anova(
  model_ex,
  ddf = "Kenward-Roger"
)
```

## Exercise 5.2: Estimate treatment means

```{r}
# Estimate marginal means
means_ex <- emmeans(
  model_ex,
  ~ ____________,
  lmer.df = "kenward-roger"
)

means_ex
```

## Exercise 5.3: Compare treatment means

```{r}
# Compare treatment means using Tukey adjustment
pairs(
  means_ex,
  adjust = "tukey"
)
```

## Exercise 5.4: Interpret the model

Answer:

1. Why is block included as a random effect?
2. What is the P-value for sulfur rate?
3. How does the mixed model differ from the one-way ANOVA?
4. Which treatment has the highest estimated mean?

# Section 6: Nonlinear Regression

For this section, sulfur rate is treated as a continuous numerical variable rather than a treatment factor.

## Exercise 6.1: Prepare the data

```{r}
# Create a numeric sulfur rate
AllData$S_rate_numeric <- as.numeric(
  as.character(AllData$S_rate_kg_ha)
)
```

## Exercise 6.2: Fit a linear-plateau model

Use grain yield as the response and sulfur rate as the explanatory variable.

```{r}
# Fit the linear-plateau model
model_LP_ex <- nls(
  ____________ ~ SSlinp(
    ____________,
    a,
    b,
    xs
  ),
  data = AllData
)

summary(model_LP_ex)
```

## Exercise 6.3: Fit a quadratic-plateau model

```{r}
# Fit the quadratic-plateau model
model_QP_ex <- nls(
  ____________ ~ SSquadp3xs(
    ____________,
    a,
    b,
    xs
  ),
  data = AllData
)

summary(model_QP_ex)
```

## Exercise 6.4: Compare the models

Use AIC to compare the linear-plateau and quadratic-plateau models.

```{r}
# Compare models using AIC
AIC(
  ____________,
  ____________
)
```

## Exercise 6.5: Plot the fitted models

Create a graph showing the observations and both fitted curves.

```{r}
# Create a sequence of sulfur rates
new_data <- data.frame(
  S_rate_numeric = seq(
    min(AllData$S_rate_numeric),
    max(AllData$S_rate_numeric),
    length.out = 50
  )
)

# Predict from the linear-plateau model
new_data$LP <- predict(
  ____________,
  newdata = new_data
)

# Predict from the quadratic-plateau model
new_data$QP <- predict(
  ____________,
  newdata = new_data
)

# Plot observations and fitted curves
ggplot() +
  geom_point(
    data = AllData,
    aes(
      x = ____________,
      y = ____________
    )
  ) +
  geom_line(
    data = new_data,
    aes(
      x = S_rate_numeric,
      y = LP,
      linetype = "Linear-Plateau"
    )
  ) +
  geom_line(
    data = new_data,
    aes(
      x = S_rate_numeric,
      y = QP,
      linetype = "Quadratic-Plateau"
    )
  ) +
  labs(
    x = "Sulfur rate (kg ha⁻¹)",
    y = "Grain yield (Mg ha⁻¹)",
    linetype = "Model"
  ) +
  theme_bw()
```

## Exercise 6.6: Interpret the results

Answer:

1. Which model has the lower AIC?
2. What is the estimated breakpoint for each model?
3. How do the fitted curves differ?
4. Which model would you use to describe the sulfur response?

