---
title: "Workshop Exercises"
subtitle: "Applied Statistics in R: From Data to Results"
format:
  html:
    self-contained: true
---

# Install Packages
```{r}
# Packages needed for the workshop
packages <- c(
  "tidyverse", "readxl", "lme4", "lmerTest",
  "emmeans", "multcomp", "nlraa"
)

# Install missing packages
install.packages(
  packages[!packages %in% rownames(installed.packages())]
)

# Load packages
lapply(packages, library, character.only = TRUE)
```


# Load packages

```{r}
# Load packages
library(tidyverse)
library(readxl)
library(lme4)
library(lmerTest)
library(emmeans)
library(multcomp)
library(nlraa)
```

# Import the data

```{r}
# Import the workshop dataset
AllData <- read_excel(
  "Data_example.xlsx",
  sheet = "data"
)
```

# Section 1: Data Exploration

## Exercise 1.1: Numerical summary

Use `summary()` to describe `grain_yield_Mg_ha`.

```{r}
# Summarize grain yield
summary(AllData$______________)
```

## Exercise 1.2: Histogram

Create a histogram of grain yield.
- `grain_yield_Mg_ha`

```{r}
# Create a histogram
ggplot(
  AllData,
  aes(x = ____________)
) +
  geom_histogram(bins = ___) +
  labs(
    x = "____________",
    y = "____________"
  ) +
  theme_bw()
```

## Exercise 1.3: Box plot

Create a box plot comparing grain yield among sulfur rates.
- `grain_yield_Mg_ha`
- `S_rate_kg_ha`
```{r}
# Create a box plot
ggplot(
  AllData,
  aes(
    x = ____________,
    y = ____________
  )
) +
  geom_boxplot() +
  labs(
    x = "____________",
    y = "____________"
  ) +
  theme_bw()
```

## Exercise 1.4: Interpret the results

Answer:

1. What is the typical grain yield?
2. How much variation is present?
3. Which sulfur rate has the highest median yield?
4. Are there any potentially unusual observations?

# Section 2: Correlation

## Exercise 2.1: Calculate a correlation

Calculate the Pearson correlation between `N_conc_R1_pct` and `S_conc_R1_pct`.

```{r}
# Calculate the Pearson correlation
cor(
  AllData$____________,
  AllData$____________,
  method = "pearson"
)
```

## Exercise 2.2: Create a scatterplot

```{r}
# Create a scatterplot
ggplot(
  AllData,
  aes(
    x = ____________,
    y = ____________
  )
) +
  geom_point() +
  labs(
    x = "____________",
    y = "____________"
  ) +
  theme_bw()
```

## Exercise 2.3: Interpret the relationship

Use your correlation coefficient and scatterplot to answer:

- Is the relationship positive or negative?
- How strong is the association?
- Does the scatterplot suggest a linear relationship?
- Are there any potentially unusual observations?

# Section 3: Linear Regression

## Exercise 3.1: Fit a linear regression

Use `N_conc_R1_pct` to predict `S_conc_R1_pct`.

```{r}
# Fit a linear regression
model_ex <- lm(
  ____________ ~ ____________,
  data = AllData
)

summary(model_ex)
```

## Exercise 3.2: Create the regression plot

```{r}
# Create a scatterplot and add the fitted regression model
ggplot(
  AllData,
  aes(
    x = ____________,
    y = ____________
  )
) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = TRUE
  ) +
  stat_poly_eq(
    aes(
      label = paste(
        after_stat(eq.label),
        after_stat(rr.label),
        sep = "*\", \"*"
      )
    ),
    formula = y ~ x,
    coef.digits = 3,
    rr.digits = 2,
    parse = TRUE,
    size = 5
  ) +
  labs(
    x = "____________",
    y = "____________"
  ) +
  theme_bw()
```

## Exercise 3.3: Interpret the model

Use the model output and graph to answer:

1.  What is the estimated slope?
2.  Is the relationship positive or negative?
3.  What does the slope mean in this context?
4.  What is the P-value for the slope?
5.  What is the $R^2$?

# Section 4: ANOVA

## Exercise 4.1: Fit the ANOVA

Use grain yield as the response and sulfur rate as the treatment.
- `grain_yield_Mg_ha`
- `S_rate_kg_ha`

```{r}
# Convert sulfur rate to a treatment factor
AllData$S_rate_kg_ha <- factor(____________)

# Fit the one-way ANOVA
model_ex <- aov(
  ____________ ~ ____________,
  data = AllData
)

summary(model_ex)
```

## Exercise 4.2: Compare treatment means

Use `emmeans()` and Tukey's method.
- `S_rate_kg_ha`

```{r}
# Estimate treatment means
emm_ex <- emmeans(
  model_ex,
  ~ ____________
)

# Compare treatment pairs
pairs(
  emm_ex,
  adjust = "tukey"
)
```

## Exercise 4.3: Interpret the results

Answer:

1. Is there evidence that grain yield differs among sulfur rates?
2. What is the ANOVA P-value?
3. Which treatment pairs differ?
4. Which sulfur rate has the highest estimated mean?

# Section 6: Nonlinear Regression

For this section, sulfur rate is treated as a continuous numerical variable rather than a treatment factor.

```{r}
# Create a numeric sulfur rate
AllData$S_rate_numeric <- as.numeric(
  as.character(AllData$S_rate_kg_ha)
)
```

## Exercise 6.1: Fit a linear-plateau model

Use grain yield as the response and sulfur rate as the explanatory variable.
- `grain_yield_Mg_ha`
- `S_rate_kg_ha`

```{r}
# Fit the linear-plateau model
model_LP_ex <- nls(
  ____________ ~ SSlinp(
    ____________,
    a,
    b,
    xs
  ),
  data = AllData
)

summary(model_LP_ex)
```

## Exercise 6.2: Fit a quadratic-plateau model

```{r}
# Fit the quadratic-plateau model
model_QP_ex <- nls(
  ____________ ~ SSquadp3xs(
    ____________,
    a,
    b,
    xs
  ),
  data = AllData
)

summary(model_QP_ex)
```

## Exercise 6.3: Compare the models

Use AIC to compare the linear-plateau and quadratic-plateau models.
- `model_LP_ex`
- `model_QP_ex`
```{r}
# Compare models using AIC
AIC(
  ____________,
  ____________
)
```

## Exercise 6.4: Interpret the results

Answer:

1. Which model has the lower AIC?
2. What is the estimated breakpoint for each model?
3. How do the fitted curves differ?
4. Which model would you use to describe the sulfur response?

